How to publish in Biological Sciences: a guide for the uninitiated

### 2023-06-15 00-author.Rmd
Add files via upload
# some corrections & additions

---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
# About the Author {- #author}

John Measey is Associate Professor of Biological Sciences at Stellenbosch University. He has authored or co-authored more than 200 peer reviewed scientific papers and book chapters, and five books. He has been the Editor-in-Chief of an ISI journal for 9 years, and currently serves as Associate Editor for 4 other journals. He has graduated more than 20 postgraduate students, and [his blog](john.measey.com/blog) on writing and publishing in biological sciences is read by thousands globally. British born and educated, he lives and works in the beautiful Western Cape, South Africa.

@@

## How to cite {-}
Measey, J. (2022) _How to publish in Biological Sciences: a guide for the uninitiated_. CRC Press, Boca Raton. 294pp. ISBN: 9781032116419  [https://doi.org/10.1201/9781003212560](https://doi.org/10.1201/9781003212560)

>>>>

# About the Author {- #author}

John Measey is Professor at the Institute of Biodiversity, Yunnan University, and Professor Extraordinary at Stellenbosch University. He has authored or co-authored [more than 200](https://orcid.org/0000-0001-9939-7615) peer reviewed scientific papers and book chapters, and five books. He has been the Editor-in-Chief of an ISI journal for 9 years, and currently serves as Associate Editor for 4 other journals. He has graduated more than 20 postgraduate students, and [his blog](john.measey.com/blog) on writing and publishing in biological sciences is read by thousands globally. British born and educated, he lives and works in the Spring City, Kunming in Yunnan Province, China and in Cape Town, South Africa.

@@

## How to cite {-}

**The printed book**

Measey, J. (2022) _How to publish in Biological Sciences: a guide for the uninitiated_. CRC Press, Boca Raton. 294pp. ISBN: 9781032116419  [https://doi.org/10.1201/9781003212560](https://doi.org/10.1201/9781003212560)

**This online version**

Measey, J. (`r format(Sys.time(), '%Y')`) _How to publish in Biological Sciences: a guide for the uninitiated_. Available from: https://howtopublishscience.org/ Accessed on: `r format(Sys.time(), '%d %B %Y')`.

### 2023-06-15 1.02-peerreview.Rmd
Add files via upload
# some corrections & additions

# What is peer review? {#peerreview}

Peer review is often considered to be the 'gold standard' of science [@mayden2012peer]. Manuscripts that have passed peer review are often considered to have been scrutinised to the highest level. If one's peers in the scientific community consider that a manuscript is worthy of publication, then it meets the high standards of peer review. In theory, the review of peers acts as the gatekeeper to all that is good in science, and excludes all that is bad. A lot has been written about peer review (>23000 articles!), and there's plenty more to read out there [@eve2021reading].

@@

This chapter provides an overview of the topic, but you may need to refer to the other chapters first depending on what your current need is.

## History of peer review
The history of peer review is surprisingly modern. We have already seen that journals themselves only date back to the 17th century (see chapter in [Part IV](#whatjournalfor)). These journals included a form of peer review in that letters concerning studies could be published, along with comments made at presentations. However, the type of systematic enforced peer review described in this book is very recent [@eve2021reading]. The journal Nature for example only started systematic peer review for its articles in 1973, and mainstream editor led peer review only really started in the late 1940s [see @tennant2017open]. Typical society journals have followed a similar form of evolution from newsletters to scholarly journals [@measey2011past].

@@

 - To assess your submission
   + Does it align with the journal?
   + Is it sound enough to send to peer review?
   + Whether to use a specialist associate editor
 - To choose the peer reviewers
   + Without potential conflicts of interest
   + Who can cover the content of the manuscript
   + Who agree to doing the review within the prescribed time-frame
 - To assess the reviews of the reviewers
   + Mitigate for potential bias in the reviews
   + Judge what is in the manuscript against what reviewers have found
   + Determine whether sufficient merit remains in order to undergo a decision (including more peer review)
 - Write the decision: When you bear in mind that the decision is likely to involve some arbitration between different reviewer opinions, and to direct the authors about what changes need to be made to a ms in order to make it acceptable, the decision is not a simple exercise. Making an editorial decision will require [careful reading of the manuscript](#editors), as well as looking past potential [biases of reviewers](#problempeerreview).

@@

### Open reviewers
Open reviewers know who the authors are, and are known by the authors and editor. Note that this is a simplification of a complex set of potential openness. For a thorough discussion see Ross-Hellauer [-@ross-hellauer2017what]. The open reviewer model does encourage good behaviour (or the avoidance of some of the worst problems) on the part of reviewers. However, reviewers remain brutally direct even when they are named, such that even open comments may be construed as bruising by the authors [@eve2021reading].

### Public reviewers {#publicreviewers}
Public reviewers know who the authors are, and are known by the authors and editor, and their names (and often their reviews) are made available to the public. This model is relatively recent, and comes along with the possibility of making the reviews with their own [DOIs](#DOI) available along with the accepted manuscript. It is worth noting that, to date, reviews for manuscripts that are rejected do not get posted using this or any other current publishing model. It does exist in the world of [preprints](#preprints2).

This could be considered the most transparent system for any journal. It has also been called Open Evaluation (OE) by Kriegeskorte et al. [-@kriegeskorte2012emerging]. [_PeerJ_](www.peerj.com) and [_eLife_](www.elifesciences.org) are among a very small handful of journals that have tried to instigate this model. Nevertheless it can be very difficult to find reviewers who are prepared to reveal their names to the authors. A study that compared _PeerJ_ publications in which reviews were made public, compared to those that authors chose to keep closed, suggested that the subsequent number of citations increased by a third for open reviews [@zong2020does]. It is only possible to speculate about why this might be. The decision at _PeerJ_ to open reviews is made first both by reviewers (who opt not to be anonymous), and then by authors (who opt to open reviews). In studies where both groups co-operate, we might hope that this results in a higher quality product. Indeed, public reviews tend to be longer, although positive comments are more frequent in closed reviews [@bornmann2012closed]. It is equally possible that authors who choose to open their peer review are more progressive and active within research (leading to more citations).

@@

I have heard some members of the community claim that they will continue to accept peer review requests until these cover what they are demanding of the community (i.e. reviews offered per year = submissions per year x ~2.5). Although this sounds very fair, I would suggest that the reality is more subtle. You shouldn't be accepting to conduct peer reviews for articles where you feel that you lack specialist knowledge. Neither should you be conducting peer review when you feel that you have a conflict of interest.

>>>>

# What is peer review? {#peerreview}
<br>
Peer review is often considered to be the 'gold standard' of science [@mayden2012peer]. Manuscripts that have passed peer review are often considered to have been scrutinised to the highest level. If one's peers in the scientific community consider that a manuscript is worthy of publication, then it meets the high standards of peer review. In theory, the review of peers acts as the gatekeeper to all that is good in science, and excludes all that is bad. A lot has been written about peer review (>23000 articles!), and there's plenty more to read out there [@eve2021reading].

@@

This chapter provides an overview of the topic, but you may need to refer to the other chapters first depending on what your current need is. 
<br>
## History of peer review
The history of peer review is surprisingly modern. We have already seen that journals themselves only date back to the 17th century (see chapter in [Part IV](#whatjournalfor)). These journals included a form of peer review in that letters concerning studies could be published, along with comments made at presentations. However, the type of systematic enforced peer review described in this book is very recent [@eve2021reading]. The journal Nature for example only started systematic peer review for its articles in 1973, and mainstream editor led peer review only really started in the late 1940s [see @tennant2017open]. Typical society journals have followed a similar form of evolution from newsletters to scholarly journals [@measey2011past].

@@

 - To assess your submission
   + Does it align with the journal?
   + Is it sound enough to send to peer review?
   + Whether to use a specialist associate editor
 - To choose the peer reviewers
   + Without potential [conflicts of interest](#conflicts)
   + Who can cover the content of the manuscript
   + Who agree to doing the review within the prescribed time-frame
 - To assess the reviews of the reviewers
   + Mitigate for potential bias in the reviews
   + Judge what is in the manuscript against what reviewers have found
   + Determine whether sufficient merit remains in order to undergo a decision (including more peer review)
 - Write the decision: When you bear in mind that the decision is likely to involve some arbitration between different reviewer opinions, and to direct the authors about what changes need to be made to a ms in order to make it acceptable, the decision is not a simple exercise. Making an editorial decision will require [careful reading of the manuscript](#editors), as well as looking past potential [biases of reviewers](#problempeerreview).

@@

### Open Peer Review {#OPR}
In Open Peer Review (OPR) authors know who reviewers are, reviewers know who authors are, and reviews alongside rebuttals are published with discrete [DOIs](#DOI) together with a published manuscript. It should be noted that manuscripts that are not accepted never have their review history published, an important flaw in the system that we will return to later. This description of Open Peer Review refers to only one (most transparent) variant, and there are at least 22 subvarients available in the literature, and a lack of common definition (for a thorough discussion see Ross-Hellauer [-@ross-hellauer2017what]). For example, some reviewers can still choose to remain anonymous in many publications (although authors cannot - for obvious reasons), and some journals force reviewers into [double-blind reviewing](#double) and then publish their comments anonymously as Open Peer Review. Other publishers allow any volunteer reviews or comments to be published alongside an article. Hence Open Peer Review generally means that reviews and comments are published alongside the article with varying degrees of reviewer anonymity. Note that it can be very difficult to find reviewers who are prepared to reveal their names to authors [see @delikoura2021open], ~40% at _PeerJ_.

A mixture of OPR and closed reviews in the same journal allows the benefits of OPR to be studied. A study that compared _PeerJ_ publications in which reviews were made public, compared to those that authors chose to keep closed, suggested that the subsequent number of citations increased by a third for OPR model [@zong2020does]. While another study found no improvement in citations for articles published in _Nature Communications_ [@ni2021influence]. Yet another study of _PLoS_ publications found a positive association between the OPR model and higher article page views, saving, sharing, and a greater HTML to PDF conversion rate, but no benefit to citations [@wei2023what]. Wei et al. [-@wei2023what] also noted that the citation index used changed whether positive benefits were found or not.

It is only possible to speculate about why more sharing and citing may be a result of OPR. The decision at _PeerJ_ to open reviews is made first both by reviewers (who opt not to be anonymous), and then by authors (who opt to open reviews). In studies where both groups co-operate, we might hope that this results in a higher quality product. Indeed, public reviews tend to be longer, although positive comments are more frequent in closed reviews [@bornmann2012closed]. It is equally possible that authors who choose to open their peer review are more progressive and active within research (leading to more citations).

OPR could be considered the most transparent system for any journal. It has also been called Open Evaluation (OE) by Kriegeskorte et al. [-@kriegeskorte2012emerging]. [_PeerJ_](www.peerj.com) PLoS and [_eLife_](www.elifesciences.org) are among a very small handful of journals that have tried to instigate this model. Interesting developments of this model have seen _PeerJ_ drop the option for authors to decline OPR as of [February 2023](https://peerj.com/about/policies-and-procedures/#open-peer-review), and _eLife_ drop the rejection of manuscripts that are reviewed as of March 2023 [@abbott2023strife], alleviating the problem noted above.

The OPR model does encourage good behaviour (or the avoidance of some of the worst problems) on the part of reviewers. However, reviewers remain brutally direct even when they are named, such that even open comments may be construed as bruising by the authors [@eve2021reading]. Another downside to OPR is that there is still the potential for [bullying](#bullying) behaviour if reviewers know who authors are when conducting a review (see [Part IV](#bullypeer)). This is then aggravated by bullies who also choose to remain anonymous, and can only be mediated by insightful editors. A solution for this might be a [double-blind reviewing](#double) model followed by Open Review Publishing of comments and rebuttal. This would encourage reviewers to interact with the publication after review and might alleviate some potential for bullying. I am not aware of this model being practised by any journal in the Biological Sciences today.

### Public reviews {#publicreviewers}
Public reviewers know who the authors are, and are known by the authors and editor, and their names (and often their reviews) are made available to the public. This model is relatively recent, and comes along with the possibility of making the reviews with their own [DOIs](#DOI) available along with the accepted manuscript. It is worth noting that, to date, reviews for manuscripts that are rejected do not get posted using this or any other current publishing model. It does exist in the world of [preprints](#preprints2).

@@

I have heard some members of the community claim that they will continue to accept peer review requests until these cover what they are demanding of the community (i.e. reviews offered per year = submissions per year x ~2.5). Although this sounds very fair, I would suggest that the reality is more subtle. You shouldn't be accepting to conduct peer reviews for articles where you feel that you lack specialist knowledge. Neither should you be conducting peer review when you feel that you have a [conflict of interest](#conflicts).

### 2023-06-15 1.04-whattowrite.Rmd
Add files via upload
# some corrections & additions

As you become more and more familiar with the academic literature you will quickly realise that actually you can publish just about anything. In this part of the book I'm going to talk about some of the most common articles that you can get published but don't feel constrained. Indeed when it comes to publishing you're only constrained by your imagination.

As an early career researcher, you will have a body of work from your thesis that you may have already published, or be in the process of publishing. These likely contain a number of data chapters that will be published as a series of papers in scientific journals. However, it's worth reflecting here about what it is possible to publish and how this might complement your existing and future publications, as well as increasing your publication portfolio with which to further your career. Certainly, having more publications is likely to increase your visibility in your community, as well as giving you more practice in academic writing.

## Standard articles
You should already be familiar with the concept of publishing standard articles and you may already have a number of these published both as a first author and as a co-author. As an Early Career Researchers, you should consider what and how you publish. For example, you should consider whether or not publishing more articles is always a good strategy? You will find a chapter that discusses this concept in detail in [Part IV](publishmore.html). In particular, you should be aware of the concept of [salami-slicing](#salami) a standard article into two or more different papers.

## Reviews
When considering the way in which citations work one of the things that you should notice is that reviews and in particular meta-analyses are cited way more times than most individual papers. For this reason one of the best things you can do is an early career researcher is to author a review on the topic of your thesis or even better a meta-analysis.

I provided an in-depth chapter on how to approach writing a review including different aspects of meta-analyses in [another chapter of this book](review.html). The importance of a timely and much used review can be seen in a lot of citation maps such as in Figure \@ref(fig:review-article). This was the first comprehensive review that sums the knowledge to that date on invasive fishes in South Africa, and so was cited most times that anyone published anything on invasive fish species in the country thereafter. Because this subject was the focus of a lot of research that happened in the area, you can see that it would logically sit at the centre of this subdiscipline. If you are unsure about whether or not a review is needed in your subdiscipline, then constructing a citation network such as that in Figure \@ref(fig:review-article) might well be useful.

@@

## Commentaries or Opinion pieces
Your opinion is important, or at least as important as anyone else's. Critical reading is a very important part of science and something that you should maintain throughout your career. From time to time you will come across articles and papers that you know are wrong. Many journals will accept commentaries or opinion pieces based on articles that they have already published. This is an opportunity for you to make a correction to something that's already published in the literature. Please know that here we are not talking about anything you think might be fraudulent for that there is another process (see here).

@@

1. If the people that wrote the article are in your network when a network close to yours then consider approaching them first about what you see as their error. You may end up getting along with them much better when you seek a solution together than writing something that antagonizes them. Even if they aren't in your network you may find a way to increase the influence of your network through soft power instead of with a commentary.

1. Check with your mentor that your interpretation of their error is correct and that pointing this out will have some value. Always try to do more than just say: "no it isn't". Many journals won't be interested in a commentary that does nothing more than show an error. If possible try and include some original data or some original analyses in your response.

1. Remember that your commentary will likely be sent immediately to the authors that you're commenting on before it is accepted by the editor. This means that they will also get a chance to comment on your commentary. However you will not get a chance to look at theirs.

@@

@@

Once all the contents of the special issue are in you can put an editoriall together to explain what the idea was of the symposium and co-authored along with your other symposium organisers. Special issues of symposium often get cited more than other articles just because it is a collection of similar things altogether in one place.

>>>>

As you become more and more familiar with the academic literature you will quickly realise that actually you can publish just about anything. In this part of the book I'm going to talk about some of the most common articles that you can get published. But don't feel constrained by what you see here. When it comes to publishing you're only constrained by your imagination.

As an early career researcher, you will have a body of work from your thesis that you may have already published, or be in the process of publishing. These likely contain a number of data chapters that will be published as a series of papers in scientific journals. However, it's worth reflecting here about what it is possible to publish and how this might complement your existing and future publications, as well as increasing your publication portfolio with which to further your career. Although publishing is not the only way to do this, having more publications is likely to increase your visibility in your community, as well as giving you more practice in academic writing.

## Standard articles
You should already be familiar with the concept of publishing standard articles and you may already have a number of these published both as a first author and as a co-author. As an Early Career Researchers, you should consider what and how you publish. For example, you should consider whether or not publishing more articles is always the right strategy for you. You will find a chapter that discusses this concept in detail in [Part IV](#publishmore). In particular, you should be aware of the concept of [salami-slicing](#salami) a standard article into two or more different papers.

## Reviews
When considering the way in which citations work one of the things that you should notice is that reviews and in particular meta-analyses are cited many more times than most individual papers. For this reason one of the best things you can do is an early career researcher is to author a review on the topic of (or around) your thesis, or even better a meta-analysis.

I provided an in-depth chapter on how to approach writing a review including different aspects of meta-analyses in [another chapter of this book](#review). The importance of a timely and much used review can be seen in a lot of citation maps such as in Figure \@ref(fig:review-article). In this example, Ellender & Weyl [-@ellender2014review] was the first comprehensive review that sums the knowledge to that date on invasive fishes in South Africa, and so was cited most times that anyone published anything on invasive fish species in the country thereafter. Because this subject was the focus of a lot of research that happened in the area, you can see that it would logically sit at the centre of this subdiscipline. If you are unsure about whether or not a review is needed in your subdiscipline, then constructing a citation network based on the key-words in your subdicipline, such as that in Figure \@ref(fig:review-article), might well be useful.

@@

## Natural History observations
Natural history notes are effectively publications made from unplanned observations. They underpin many of the [Big Ideas](https://howtowriteaphd.org/bigidea.html) in biological sciences, including the theory of evolution [@darwin1859origin]. For an insightful read on the interaction of ecology and natural history, see Anderson [-@anderson2017why].

Natural history observations are accounts of novel information, once-off or multiple/methodical observations and range from a full paper to a few structured paragraphs. They are defined by  not being the outcome of a planned experiment or inquiry. That is, they would not be part of your [research proposal](https://howtowriteaphd.org/proposal.html) or [preregistered plan](https://howtowriteaphd.org/transparency.html#preregistration) for your research. This does not make them any less important than one of your planned studies, but statistically they generally have less power and could potentially be victim of [HARKing](https://howtowriteaphd.org/nohypothesis.html#avoid-harking). As a result, it is important to clearly state in the Methods that they are natural history observations, so that others know how to interpret them.

Important natural history observations require replication as planned experiments to determine their validity and hence importance moving forwards. They are a good example of how [preregistered plan](https://howtowriteaphd.org/transparency.html#preregistration) for your research do not overly restrict publishing your work, but do keep you honest in your reporting.

## Commentaries or Opinion pieces {#commentary}
Your opinion is important, or at least as important as anyone else's. Critical reading is a very important part of science and something that you should maintain throughout your career. From time to time you will come across articles and papers that you know are wrong, or fail to represent sufficient balance. Many journals will accept commentaries (also known as rebuttals) or opinion pieces based on articles that they have already published. This is an opportunity for you to make a correction to something that's already published in the literature. Please know that here we are not talking about anything you think might be fraudulent for that there is [another process](#fraud).

@@

1. If the people that wrote the article are in your network when a network close to yours then consider approaching them first about what you see as their error. You may end up getting along with them much better when you seek a solution together than writing something that antagonises them. Even if they aren't in your network you may find a way to increase the influence of your network through soft power instead of with a commentary.

1. A commentary should never be an _ad hominem_ attack. Never comment on the authors, only their content.

1. Check with your [mentor](#mentor) or another colleague that your interpretation of their error is correct and that pointing this out will have some value. Always try to do more than just say: "no it isn't". Many journals won't be interested in a commentary that does nothing more than show an error. If possible try and include some original data or some original analyses in your response.

1. Remember that your commentary will likely be sent immediately to the authors that you're commenting on before it is accepted by the editor. This means that they will also get a chance to comment on your commentary. However you will not get a chance to look at their comment.

@@

1. Don't expect that your rebuttal will change the way that people think. A study of seven prominent papers and their rebuttals showed that rebuttals were rarely cited alongside the prominent paper, and most citations were not critical [@banobi2011rebuttals].

@@

Once all the contents of the special issue are in you can put an editorial together to explain what the idea was of the symposium and co-authored along with your other symposium organisers. Special issues of symposium often get cited more than other articles just because it is a collection of similar things altogether in one place.

### 2023-06-15 2.03-suggestreviewers.Rmd
Add files via upload
# some corrections & additions

 - Anyone who is an author or in the acknowledgements
 - Anyone else you feel may have a conflict of interest 
   + connected with the work but not an author
   + someone who was on your review committee
 - Someone in your department or even at your institution (there may be exceptions here, but often journals will not consider people with matching institutional emails)
 - friends, labmates or even relatives (even if you genuinely think that they would do a good job)
 - people who are regular co-authors (grant panels specifically ban you from naming these people as potential reviewers. If you have a large network, this can be problematic)

@@

In addition to allowing you to suggest reviewers, many journals allow you to oppose reviewers. I usually leave this blank. If you have a particular lab or persons who you know will not follow the [spirit of peer review](#spirit), you could enter them here.

## Who will the editor use?
It is worth bearing in mind that whoever you suggest, the editor may not use them. I have talked to editors who will never consider anyone recommended by authors as a matter of course, because they assume that these people will be positively biased towards the authors. This same editor said that they'd always use at least one of the opposed reviewers. This rarely fits with journal policy.

>>>>

 - Anyone who is an author or in the acknowledgements
 - Anyone else you feel may have a [conflict of interest](#conflicts) 
   + connected with the work but not an author
   + someone who was on your review committee
 - Someone in your department or even at your institution (there may be exceptions here, but often journals will not consider people with matching institutional emails)
 - friends, labmates or even relatives (even if you genuinely think that they would do a good job)
 - people who are regular co-authors (grant panels specifically ban you from naming these people as potential reviewers. If you have a large network, this can be problematic)

@@

In addition to allowing you to suggest reviewers, many journals allow you to oppose reviewers. Most people leave this blank [@fox2017author-suggested]. If you have a particular lab or persons who you know will not follow the [spirit of peer review](#spirit), you could enter them here.

## Who will the editor use?

It is worth bearing in mind that whoever you suggest, the editor may not use them. I have talked to editors who will never consider anyone recommended by authors as a matter of course, because they assume that these people will be positively biased towards the authors. Other editors say that they always use at least one of the opposed reviewers. Such extreme editors do exist, but most appear to fall somewhere in between.

Studies show that author suggested reviewers are less likely to reject a manuscript in the initial stages of peer review than those chosen by the editorial team, even though the quality and tone between the two sets of reviewers are no different [@wager2006are; @schroter2006differences; @rivara2007comparison; @fox2017author-suggested; @liang2018should]. According to Fox et al. [-@fox2017author-suggested], for the journal _Functional Ecology_, editors approach author proposed reviewers about a quarter of the time. Because proposed reviewers are more likely to accept invitations [@liang2018should], the likelihood of them reviewing your manuscript is higher.

Editorial policy will dictate whether or not opposed reviewers are used, but analyses in some biological journals report that opposed reviewers are occasionally used [@fox2017author-suggested; @hausmann2018reviewer]. These differences in editorial policy may explain why some journals find that opposing reviewers increases the chances of acceptance [@grimm2005suggesting].

### 2023-06-15 4.07-retract.Rmd
Add files via upload
# some corrections & additions

### Fraud by individuals or laboratories
The fabrication of data does happen. An anonymous survey on research integrity in the Netherlands suggested that prevalence of fabrication was 4.3% and falsification was 4.2%, while nearly half of those surveyed admitted to questionable research practices most prevalent in PhD candidates and junior researchers [@gopalakrishna2021prevalence]. A growing body of retractions and alleged evidence on the tampering of data in spreadsheets has led to the suspension of a top Canadian researcher, [Jonathan Pruitt](#Pruittdata). The detection of fraudulent (usually duplicated) data in spreadsheets is not too difficult to spot (e.g. by application of Benford's law - the last digits of naturally occurring numbers should approach uniformity), and has become the subject of some contract data scientists who specialise in finding such instances of fraud. To some extent, the automated assessment of fraudulent practices has been or could be implemented for many infringements [@bordewijk2021methods].

@@

### By third parties

The existence of [paper mills](#papermills) should also be mentioned at this point. Papers, produced commercially and to order, are entirely fabricated by third parties to improve the CVs of real paying scientists [@teixeiradasilva2021abuse]. Hundreds of these papers have been discovered that appear to come from the same source [@bik2020tadpole], but the true size of such additions to the scientific literature is unknown.

@@

If the research is published, and you think it is fraudulent, approach the editor directly. If there is some conflict of interest (like the person is at your institution), then you can try to sort it out internally (as described above). Otherwise, you can approach the editor directly yourself, anonymously or by using a third party.

@@

 -  ['Image manipulation in a published article'](https://publicationethics.org/resources/flowcharts/image-manipulation-published-article) [@cope2018what]
 -  ['How to recognise potential authorship problems'](https://publicationethics.org/resources/flowcharts/how-recognise-potential-authorship-problems) [@cope2018how]
 -  ['Systematic manipulation of the publication process'](https://publicationethics.org/resources/flowcharts/systematic-manipulation-publication-process) [@cope2018systematic]
 - ['How to recognise potential manipulation of the peer review process'](https://publicationethics.org/resources/flowcharts/how-recognise-potential-manipulation-peer-review-process) [@cope2017how]
 - ['If you suspect fabricated data in a submitted manuscript'](https://publicationethics.org/resources/flowcharts-new/what-do-if-you-suspect-fabricated-data-0) [@wager2006suspecteda]
 - ['Ghost, guest, or gift authorship in a submitted manuscript'](https://publicationethics.org/resources/flowcharts-new/changes-authorship) [@wager2006suspectedb]
 - Undisclosed conflict of interest 
   + [in a submitted manuscript](https://publicationethics.org/resources/flowcharts-new/conflict-interest)
   + [in a published article](https://publicationethics.org/resources/flowcharts/undisclosed-conflict-interest-published-article)
 - ['Plagiarism in a submitted manuscript'](https://publicationethics.org/node/19701) [@wager2006suspected]

## Confirmation Bias and the paradox of high-flying academic careers {#PruittData}
Jonathan Pruitt had it all going for him. His studies of spider sociality were producing novel and significant results that opened the door to publications in high impact journals. In turn, this opened the door to getting prizes and funding. The funding allowed him to conduct more studies and soon a prestigious chair in Canada with more funding to pursue his rocketing career. Things started to unravel for Pruitt when colleagues raised concern about the data in some of his publications. Things gathered pace very quickly, and doubt gathered around more and more of his publications. Although there is much written about the Pruitt debacle on the internet, the blog by *American Naturalist* editor and former Pruitt fan and friend, Dan Bolnick, is particularly enlightening [@bolnick202117]. Pruitt's case is becoming increasingly untenable as more editors backed by co-authors are retracting papers where he contributed data [see @marcus2020spider]. For Pruitt, this has become a threat to his career and livelihood [@pennisi2020embattled]. Similarly, his university is facing the possibility that they hired a fraud. Consequently, this whole debacle has slipped into the legal world. Bolnick has clearly suffered personally from the affair, but has set out to provide as transparent an account as possible.

The harrowing part of Bolnick's account is when he, co-authors and other editors started to receive letters from Pruitt's Lawyer (see one example on [Bolnick's blog](https://ecoevoevoeco.blogspot.com/2021/05/17-months.html)). At the point that the lawyer steps in, the functioning academic community that had raised itself to meet the demands of the concerns began to get muted. Bolnick then makes an important point that the legal threats from Pruitt's Lawyer were stifling the freedom for academics (in this case the co-authors and editors) to publish, and therefore their [academic freedom](#tenure).

>>>>

### Fraud by individuals or laboratories
The fabrication of data does happen. An anonymous survey on research integrity in the Netherlands suggested that prevalence of fabrication was 4.3% and falsification was 4.2%, while nearly half of those surveyed admitted to questionable research practices most prevalent in PhD candidates and junior researchers [@gopalakrishna2021prevalence]. A growing body of retractions and alleged evidence on the tampering of data in spreadsheets has led to the suspension, resignation and ultimately publication of the misdeeds of a top researcher, [Jonathan Pruitt](#Pruittdata). The detection of fraudulent (usually duplicated) data in spreadsheets is not too difficult to spot (e.g. by application of Benford's law - the last digits of naturally occurring numbers should approach uniformity), and has become the subject of some contract data scientists who specialise in finding such instances of fraud. To some extent, the automated assessment of fraudulent practices has been or could be implemented for many infringements [@bordewijk2021methods].

@@

### By third parties {#3rdparties}

The existence of [paper mills](#papermills) should also be mentioned at this point. Papers, produced commercially and to order, are entirely fabricated by third parties to improve the CVs of real paying scientists [@teixeiradasilva2021abuse]. Hundreds of these papers have been discovered that appear to come from the same source [@bik2020tadpole], but the true size of such additions to the scientific literature is unknown. For example, in April 2023 Wiley admitted that their own publications (under [Hindawi](#Hindawi)) had ~1700 papers that resulted from paper mills. This will undoubtedly be another bump in the number of retractions as Wiley struggle to sniff out all of the faked content [@flynn2023guest].

@@

If the research is published, and you think it is fraudulent, approach the editor directly. If there is some [conflict of interest](#conflicts) (like the person is at your institution), then you can try to sort it out internally (as described above). Otherwise, you can approach the editor directly yourself, anonymously or by using a third party.

@@

 -  ['Image manipulation in a published article'](https://publicationethics.org/resources/flowcharts/image-manipulation-published-article) [@cope2018what]
 -  ['How to recognise potential authorship problems'](https://publicationethics.org/resources/flowcharts/how-recognise-potential-authorship-problems) [@cope2018how]
 -  ['Systematic manipulation of the publication process'](https://publicationethics.org/resources/flowcharts/systematic-manipulation-publication-process) [@cope2018systematic]
 - ['How to recognise potential manipulation of the peer review process'](https://publicationethics.org/resources/flowcharts/how-recognise-potential-manipulation-peer-review-process) [@cope2017how]
 - ['If you suspect fabricated data in a submitted manuscript'](https://publicationethics.org/resources/flowcharts-new/what-do-if-you-suspect-fabricated-data-0) [@wager2006suspecteda]
 - ['Ghost, guest, or gift authorship in a submitted manuscript'](https://publicationethics.org/resources/flowcharts-new/changes-authorship) [@wager2006suspectedb]
 - Undisclosed [conflict of interest](#conflicts) 
   + [in a submitted manuscript](https://publicationethics.org/resources/flowcharts-new/conflict-interest)
   + [in a published article](https://publicationethics.org/resources/flowcharts/undisclosed-conflict-interest-published-article)
 - ['Plagiarism in a submitted manuscript'](https://publicationethics.org/node/19701) [@wager2006suspected]

## Confirmation Bias and the paradox of high-flying academic careers {#PruittData}
Jonathan Pruitt had it all going for him. His studies of spider sociality were producing novel and significant results that opened the door to publications in high impact journals. In turn, this opened the door to getting prizes and funding. The funding allowed him to conduct more studies and soon a prestigious chair in Canada with more funding to pursue his rocketing career. Things started to unravel for Pruitt when colleagues raised concern about his data in some of their publications. Things gathered pace very quickly, and doubt gathered around more and more of his publications. Although there is much written about the Pruitt debacle on the internet, the blog by *American Naturalist* editor and former Pruitt fan and friend, Dan Bolnick, is particularly enlightening [@bolnick202117]. Pruitt's case is becoming increasingly untenable as more editors backed by co-authors are retracting papers where he contributed data [see @marcus2020spider]. For Pruitt, this became a threat to his career and livelihood [@pennisi2020embattled]. Similarly, his university is faced the possibility that they hired a fraud. Consequently, this whole debacle has slipped into the legal world. After three and a half years, McMaster University (his former employer) has finally issued [their report](https://secretariat.mcmaster.ca/app/uploads/RI-Pruitt-summary-of-findings.pdf) on the affair concluding that he engaged in falsification and fabrication of data [@kincaid2023spider]. Pruitt is thought to have left research and taken a teaching job at a high school in Florida [@kincaid2023spider].

The harrowing part of Bolnick's account is when he, co-authors and other editors started to receive letters from Pruitt's Lawyer (see one example on [Bolnick's blog](https://ecoevoevoeco.blogspot.com/2021/05/17-months.html)). At the point that the lawyer steps in, the functioning academic community that had raised itself to meet the demands of the concerns began to get muted. Bolnick then makes an important point that the legal threats from Pruitt's Lawyer were stifling the freedom for academics (in this case the co-authors and editors) to publish, and therefore their [academic freedom](#tenure). Although there are many lessons from the Pruitt debacle, the important points include the impact that it had on other people's lives and careers, as well as the legal implications of raising grant monies on the basis of fabricated data. Although there might be some comfort in the way in which the academic community pulled together and that the university inquiry ultimately reached the right decision, there are countless other cases of academic fraud and misrepresentation where the perpetrators are never brought to justice.
